Exploring the Presidents Corpus

### 2019-10-18 Source/Paper 1 - Presidents Corpus Paper/PresidentsCorpus.tex
updated pres paper
updated changes from HV

Alzheimer's Disease (AD) and other forms of dementia affect a significant proportion of the global older adult population and the impact on morbidity and mortality rates is considerable. Dementia and AD are currently the leading cause of death in England and Wales and the sixth leading cause of death in the United States (US). A recent report commissioned by the Alzheimer's Society in 2015 estimated that in the UK by 2015 there would be approximately 855,000 people rising to 1 million by 2021. This represents 1 in 79 of the general population rising to 1 in 14 of those aged 65 or over \cite{AlzheimersSociety2014}. Worldwide, there are 46 million people with a diagnosis of dementia globally and that number is expected to hit 131.5 million by 2050 \cite{Prince2015}. From a financial perspective, they estimated annual spend on dementia healthcare is £4.3 billion of which approximately £85 million is spent on diagnosis. The total financial burden of dementia (excluding the costs associated with early onset dementia) is £26.3 billion annually. Globally, this picture is a lot bleaker. Another report by Alzheimer's Disease International suggests that in 2015 there were  The report also states that the worldwide cost of AD in 2018 is estimated to be in the region of one trillon US dollars.

@@

AD is a neurodegenerative disease in which a definitive diagnosis can only be produced at post-mortem. However, there are a number of psychological and physiological indicators that can indicate that dementia is present. From a physiological perspective, researchers have identified two proteins called beta-amyloid and tau. In a typical case, tau accumulates and eventually forms tangles inside neurons and beta-amyloid clumps into plaques which slowly builds up between neurons. At a certain point, the levels of beta-amyloid rise and trigger a more rapid spread of tau throughout the brain. Eventually, due to this and other changes, neurons lose their ability to communicate and the brain starts to shrink. This leads to some of the more psychological symptoms, those who have dementia demonstrate cognitive deficits such as problems with episodic and semantic memory, organizing and planning, difficulties with language, problems with executive function and visuospatial deficits \cite{McKhann2011}. In addition, these symptoms are often accompanied by emotional problems such as depression and behavioural difficulties. As more neurons die throughout the brain, a person with Alzheimer's gradually loses the ability to think, remember, make decisions and function independently.

@@

Despite this growing problem and an increasing understanding about how AD affects the brain there are no medications that improve the prognosis of those with AD. All the medications that are currently on the market are designed to manage symptoms. Whilst there are numerous investigational drugs in development for the treatment of AD, a larger than normal percentage (99.6\%) of these drugs fail in clinical trials (in contrast to anti-cancer drugs which have a 80\% failure rate) \cite{Cummings2014}. Researchers have proposed that a possible reason for the lack of success is that the drugs treatments are initiated too far along in the progression of the disease and thus much of the degeneration of the brain has already taken place \cite{Cummings2014}. Research has started to be more focused on AD at it's earliest stages which some literature describes as 'Mild Cognitive Impairment (MCI) due to AD'.

@@

One of the challenges of this approach is differentiating natural cognitive decline due to aging with decline due to a form of cognitive impairment or dementia. This challenge is often complicated further due to the large variation in the cognitive abilities and educational background of individuals. Albert and his team have worked to define clinical criteria which professionals can use to diagnose MCI due to AD and differentiate this from age-associated memory impairment and age-associated cognitive decline.  One of the most important observations from this piece of work is that a diagnosis of MCI requires evidence of intra-individual change and optimally requires evaluation at two or more points \cite{Albert2011}, and this is essentially to place more importance on the trajectory of a person's cognitive abilities rather than a person's cognitive ability in general. 

@@

There has been a significant research in the area of language deterioration as a means of detecting Alzheimer's Disease. This usually takes the form analysis of speech recorded as part of a cognitive assessment such as the Picture Description Task \cite{Orimaye2014,Fraser2015}. Given that language samples are relatively easy to collect, research has moved towards analysis of spontaneous speech. An good example of this type of research is the study conducted by Berisha and Liss which looked at the differences in language use between two US presidents, Ronald Reagan (RR), who would go on to receive a diagnosis of Dementia and George H. W. Bush (GHWB) who acted as a matched control based on Age \cite{Berisha2015}. They found several differences in language use over time which they felt acted as indicators of RR's difficulties with language due to dementia. These significant differences were in the number of unique words used per speech, the use of non-specific nouns and fillers and low-imageability verbs \cite{Berisha2015}. 

@@

This study replicates work done by Berisha and Liss and extends this by adding Donald Trump (DJT) as an alternative, more appropriate comparison to RR as he is much closer in age than GHWB. This work contributes to our understanding of the pre-diagnosis phase of Alzheimer's Disease and it's effect on language. We explore a wider range of features than previously looked at with this dataset and our hypothesis is that there will be features of RR's language which change more noticably over time when compared with the two closest appropriate controls. In addition, our assumption is that any language deterioration found will be non-linear and so we will be looking at non linear and linear correlations within the data to see if decline is constant over time

\section{Methodology}\label{methodology}
We took 46 transcripts of Ronald Reagan’s (RR) press conferences from 1981 to 1988 and compared them with 134 press conferences by George H. W. Bush (GHWB) and 29 press conferences conducted by Donald J. Trump (DJT).  We analyzed transcripts for lexical features shown to change longitudinally with dementia. These features encompassed, word level, sentence level and document level features and included a number of features contained in the study by Berisha and Liss with the aim of replicating and extending on their findings. We compared the trends described in the transcripts of RR and GHWB, but also included DJT. Berisha and Liss originally made the comparison as it GHWB (GHWB - age at the start of presidency - 64 years, 222 days) was the closest match to RR in terms of age (RR  - age at the start of presidency, 69 yeaars and 349 days). However, with the inauguration of Trump, he now is the closest comparible president in terms of age (DJT - age at the start of presidency - 70 years, 220 days). It would be interesting to look at a comparison of RR and DJT to see whether the comparisons made by Berisha and Liss hold true with this more appropriate match (in terms of age). DJT as with GHWB has no known diagnosis of AD. We used the press conference transcripts in the American Presidency Project (APP) archive as a data source for this project. The APP is a comprehensive and organized searchable database of presidential documents, including transcripts of speeches, transcripts of news conferences, and other public documents.

@@

\begin{figure}[H]
\centering
\includegraphics[width=300px, height=300px]{RRppromLM.jpg}
\caption{RR Personal Pronouns - Linear Model}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=300px, height=300px]{RRppromLM.jpg}
\caption{RR Personal Pronouns - Local Regression Model}
\end{figure}

@@

\bibliographystyle{unsrt}
\bibliography{main}

>>>>

The diagnosis of dementia is generally made when there is a decline in brain function due to physical changes in the brain \cite{Albert2011}. It affects a significant proportion of the global older adult population and the impact on morbidity and mortality rates is considerable. Dementia is currently the leading cause of death in England and Wales and the sixth leading cause of death in the United States (US). A 2014 report commissioned by the Alzheimer's Society estimated that in the UK by 2015 there would be approximately 855,000 people rising to 1 million by 2021 \cite{}. This represents 1 in 79 of the total UK population rising to 1 in 14 of those aged 65 or over \cite{AlzheimersSociety2014}. Worldwide, there are 46 million people with a diagnosis of dementia globally and that number is expected to hit 131.5 million by 2050 \cite{Prince2015}. From a financial perspective, the cost burden is also significant. The estimated annual spend on dementia healthcare in the UK is £4.3 billion of which approximately £85 million is spent on diagnosis. The total financial burden of dementia (excluding the costs associated with early onset dementia) is £26.3 billion annually. Globally, this picture is a lot bleaker. Another report by Alzheimer's Disease International suggests that in 2015 there were  The report also states that the worldwide cost of AD in 2018 is estimated to be in the region of one trillon US dollars \cite{}.

@@

There are different types of dementia including Alzheimer's Disease (AD), vascular dementia, dementia with Lewy bodies and fronto-temporal dementia, all of which are currently incurable. AD is a progressive neurodegenerative disease and is the most common type of dementia, responsible for approximately XX\% of all cases \cite{}. Currently, a definitive diagnosis for AD can only be produced at post-mortem. However, there are a number of psychological and physiological indicators that can indicate that dementia is present. From a physiological perspective, researchers have identified two proteins called beta-amyloid and tau have been identified as two of the key proteins that have a role in initiation and progression of AD. In a typical case beta-amyloid clumps into plaques which slowly build up between neurons with tau proteins accumulating and eventually forming tangles inside neurons. At a certain point, the levels of beta-amyloid rise triggering a more rapid spread of tau throughout the brain. Eventually, due to these and other changes, neurons lose their ability to communicate and the brain starts to shrink. This leads to the development of some of the psychological symptoms associated with AD, primarily cognitive deficits such as problems with episodic and semantic memory, organizing and planning, difficulties with language, problems with executive function and visuospatial deficits \cite{McKhann2011}. These cognitive symptoms are often accompanied by emotional problems such as depression and behavioural difficulties. As more neurons die throughout the brain, a person with AD gradually loses the ability to think, remember, make decisions and function independently.

@@

Despite the increasing prevalence of AD and an improved understanding about how it affects the brain there are no medications that improve prognosis. All the medications that are currently on the market are designed to manage symptoms. Whilst there are numerous investigational drugs in development for the treatment of AD, a larger than normal percentage of these drugs fail in clinical trial stage of the drug discovery process ((99.6\% failure rate vs 80\% for systemic anti-cancer drugs) \cite{Cummings2014}. Cummings et al proposed that a possible reason for the lack of success is that the drugs treatments are initiated too far along in the progression of the disease and thus much of the degeneration of the brain has already occurred \cite{Cummings2014}. Research focus has now started to shift to the earlier stages of AD (i.e. symptomatic pre-dementia phase of AD) which some literature describes as 'Mild Cognitive Impairment (MCI) due to AD'.

@@

One of the challenges associated with the early detection of AD is differentiating natural age associated memory impairment and cognitive decline due to aging from with decline due to AD. This challenge is often complicated further due to the large variation in the cognitive abilities and educational background of individuals. The work of Albert et al helps to address this by the development of clinical criteria which professionals can use to diagnose MCI due to AD. One of the most important observations from this piece of work is that a diagnosis of MCI requires evidence of intra-individual change and optimally requires evaluation at two or more points \cite{Albert2011}, and this is essentially to place more importance on the trajectory of a person's cognitive abilities rather than a person's cognitive ability in general. 

@@

There has been a significant research in the area of language deterioration as a method of detecting AD at an earlier stage. This usually takes the form of recording speech whilst patients' undertake a cognitive assessment such as the Picture Description Task \cite{Orimaye2014,Fraser2015}. Given that language samples are relatively easy to collect, research has moved towards analysis of spontaneous speech. The work of Berisha and Liss is a good example of this, this study compared the differences in language use between two US presidents, Ronald Reagan (RR), who would go on to receive a diagnosis of Dementia and George H. W. Bush (GHWB) who acted as a matched control based on Age \cite{Berisha2015}. They found several differences in language use over time which they felt acted as indicators of RR's difficulties with language due to AD. The statistically significant differences identified included the number of unique words used per speech, the use of non-specific nouns and fillers and low imageability verbs \cite{Berisha2015}. 

@@

This study replicates the research of Berisha and Liss and extends this by adding Donald Trump (DJT) as an additional comparison to RR. This study aims to contribute to our understanding of the pre-diagnosis phase of Alzheimer's Disease and it's effect on language. Our hypothesis is that there will be features of RR's language which change more noticably over time when compared with the two closest appropriate controls. Compared to the previous research study, we explored a wider range of features within this dataset. In addition, our assumption was that any language deterioration found would be non-linear and so we will be looking at non linear and linear correlations within the data were looked at to see if language decline is constant over time

\section{Methodology}\label{methodology}
We took 46 transcripts of press conferences given by RR from 1981 to 1988 and compared them with 134 press conferences from 1989 to 1993 given by GHWB and 29 press conferences from 2016 to 2019 conducted by DJT.  We analyzed transcripts for lexical markers (described below) shown to change longitudinally with AD. These lexical features encompassed, word level, sentence level and document level features and included a number of features contained in the study by Berisha and Liss. 
\par 
In the original study, GHWB was selected as the comparator president as he was the closest match in terms of age to RR (GHWB - age at the start of presidency: 64 years and 222 days vs RR: 69 years and 349 days). However, since his inauguration, DJT is now the closest comparible president in terms of age (DJT - age at the start of presidency: 70 years and 220 days). We included DJT who like GHWB has no known diagnosis of AD to determine whether the comparisons made by Berisha and Liss hold true with this closer presidential match in terms of age. 
\par 
We used the press conference transcripts in the American Presidency Project (APP) archive as a data source for this project. The APP is a comprehensive and organized searchable database of presidential documents, including transcripts of speeches, transcripts of news conferences, and other public documents. These documents are open access and can be downloaded at any time from the APP archive.

@@

\begin{figure}[b]
	\centering
	\begin{minipage}[b]{0.4\textwidth}
		\includegraphics[width=150px, height=150px,  natwidth=300, natheight=300]{RRppronLM.jpg}
		\caption{RR Personal Pronouns - Linear Model}
	\end{minipage}

	\begin{minipage}[b]{0.4\textwidth}
		\includegraphics[width=150px, height=150px, natwidth=300, natheight=300]{RRppronLOESS.jpg}
		\caption{RR Personal Pronouns - Local Regression Model}
	\end{minipage}		
\end{figure}

@@

\bibliographystyle{unsrt}
\bibliography{main3}
